Refinement for open automata

### 2021-04-06 2021-Refinement/report.tex
Refinement v0something
	new file:   preamble.tex
	renamed:    rapport.pdf -> report.pdf
	renamed:    rapport.tex -> report.tex

\theoremstyle{plain}
\newtheorem{thm}{Theorem}
\newtheorem{lem}{Lemma}
\newtheorem{cor}{Corollary}
\newtheorem{prop}{Proposition}

\theoremstyle{definition}
\newtheorem{defi}{Definition}
\newtheorem{exi}{Example}

\newcommand\comment[3]{\colorbox{#1}{#2}\marginpar{#3}}
\newcommand\Rabea{\comment{yellow}}
\newcommand\Ludo{\comment{green}}
\newcommand\Eric{\comment{cyan}}
\newcommand\Quentin{\comment{pink}}

\newcommand\nmm[1]{\(\displaystyle #1\)} % nmm for Nice Math Mode
\newcommand\setR{\mathbb{R}}
\newcommand\setZ{\mathbb{Z}}
\newcommand\setN{\mathbb{N}}
\newcommand\mpar[1]{{\left(#1\right)}}
\newcommand\mbrk[1]{{\left[#1\right]}}
\newcommand\mbrc[1]{{\left\{#1\right\}}}
\newcommand\mdbrk[1]{\left⟦#1\right⟧}
\newcommand\card[1]{{\left|#1\right|}}
\newcommand\psubst[1]{\mbrc{\!\!\mbrc{#1}\!\!}}
\newcommand\subst[2]{\mbrk{#1\middle/#2}}
\newcommand\choice[1]{\left\{\everymath{\displaystyle}%
	\begin{array}{lr}#1\end{array}\right.}
\newcommand\midbar{\,\middle|\,}
\newcommand\mset[2]{\mbrc{#1\midbar #2}}
\newcommand\subbox[1]{{\makebox[.5\width]{\(\scriptstyle #1\)}}}
\newcommand\bigsymb[2][\Large]{\text{#1\nmm{#2}}} % DeclareMathDelimiter?
\newcommand\defnotation{\DOTSB\;{\Colon=}\;}
\newcommand\defobject{\DOTSB\;{\coloneq}\;}
\newcommand\nwedge{\DOTSB\;{\wedge}\;} % when you want to force space
\newcommand\qwedge{\DOTSB\quad{\wedge}\quad}
\newcommand\fvars[1]{{\mathit{vars}\mpar{#1}}}
\newcommand\terms{{\mathcal{T}}}
\newcommand\cterms{{\mathcal{T}_{\makebox[1pt][r]{\(\scriptstyle \emptyset\)}}}}
\newcommand\formulae{{\mathcal{F}}}
\newcommand\cformulae{{\mathcal{F}_{\makebox[1pt][r]{\(\scriptstyle \emptyset\)}}}}
\newcommand\values{{\mathcal{P}}}
\newcommand\OA[6]{\left<#1,#2,#3,#4,#5,#6\right>}
\newcommand\OT[6]{\text{\small\(%
	\setlength\arraycolsep{2pt}\everymath{\displaystyle}%
	\renewcommand\arraystretch{1}%
	\begin{array}{c}%
	#4,#5,#6 \\\noalign{\color{red}\hrule height 1pt}
	#1\xrightarrow{\raisebox{-1.5pt}[.8\height][0pt]{\makebox[1.4\width]{\(\scriptstyle #3\)}}}#2
	\end{array}\)}}
\newcommand\OTx[4]{\OT{s_{#1}}{s'_{#1 #2}}{\alpha_{#1 #2}}{\beta_{#3 j}^{j \in J'_{#4}}}{g_{#1 #2}}{\psi_{#1 #2}}}
\newcommand\OTg{\OTx{}{}{}{}}

@@

Family of values, or equivalently maps will be noted \(\mset{i \mapsto x_i}{i \in I}\), \(\mset{i \gets x_i}{i \in I}\) or \(x_i^{i \in I}\).
The latter will only be used when unambiguous; for instance \(\mpar{ax}^{x \in \setR}\) represents a scaling function, \(c^{i \in I}\) is a constant function over \(I\), but \(\mbrc{\alpha \mapsto 1, \beta \mapsto 2, \gamma \mapsto 3}\) must be represented with one of the two first notation.
The disjoint union of two maps \(\varphi: I \to X\) and \(\psi: J \to Y\) with \(I \cap J = \emptyset\) is \(\varphi \uplus \psi: \mpar{I \uplus J} \to \mpar{X \cup Y}\).

@@

\begin{defi}[Free variables, Closed terms, Closed formulae]
\(\fvars{e}\) is the set of unbound variables in \(e \in E\).
The terms restricted to variables in \(V\) are \(\terms_V \defnotation \mset{t \in \terms}{\fvars{t} = \emptyset}\); \(\terms_V \subset \terms\).
The formulae restricted to variables in \(V\) are \(\formulae_V \defnotation \mset{f \in \formulae}{\fvars{f} = \emptyset}\); \(\formulae_V \subset \terms\).
The closed terms are terms restricted to variables in \(\emptyset\), \(\cterms\).
The closed formulae are formulae restricted to variables in \(\emptyset\), \(\cformulae\).
\end{defi}
By extension, expressions restricted to variables and closed expressions are also defined.
A closed expression is an expression without unbound variable.
\begin{defi}[Interpretation, Values, Satisfiability, (Parallel) substitution]
We assume that the following functions are given:
\begin{itemize}
\item The interpretations of closed terms, \(\mdbrk{\_}: \cterms \to \values\), with \(\values\) a set of interpreted value of terms;
\item The satisfiability relation on closed formulae, \({\vdash} \subseteq \cformulae\) and unsatisfiability relation \({\nvdash} \subseteq \cformulae\);
\item The substitution in \(e \in E\) of \(x \in \fvars{e}\) by \(t \in \terms\), \(e\subst{t}{x}\);
\item The parallel substitution in \(e \in E\) of variables in \(V\) by \(\psi: V \to \terms\), \(e\psubst{\psi}\).
\end{itemize}
\end{defi}
For the parallel substitution, the set \(V\) is not required to be a subset of \(\fvars{e}\).
In the case where it is not, the variables in \(V \setminus \fvars{e}\) are not substituted.
The substitutions might give a nonsensical expression; for instance terms can be integers and pairs with addition, \(\mpar{a + b}\psubst{a \mapsto 7, b \mapsto \mpar{4, 5}}\) is not a valid term.
This can be guarded with the atom \(e\subst{t}{x} \in E\) and \(e\psubst{\psi} \in E\) and it will implicitely be the case for instance when there is quantification on \(t\) and \(\psi\) to simplify notations.

The interpretation of values is supposed to be decidable.
The (un)satisfiability of formulae might not be decidable nor complete nor consistent, however we will pretend like they are because these are really hard problems for logicians that we don't want to deal with.\footnote{In practise a term algera with only equality is used, and it is has all these properties.}

A value \(v\) may be used for keeping a variable state, and then injected in terms for substitution.
While this is correct when \(\values \subseteq \cterms\), in the other cases this will be used as a shorthand for substitution with any \(t \in \mdbrk{v}^{-1}\).

We are not interested in how \(\values\) is defined.
Whether it is predefined or caracterised by \(\values \defnotation \mdbrk{\cterms}\) is not the point of this definition.

The satisfiability relations will be noted using the classical relational notation \(\vdash f \defnotation f \in {\vdash}\) and \(\nvdash f \defnotation f \in {\nvdash}\).
The usual satisfiability under a partial valuation is noted \nmm{\sigma \vdash f \defnotation \exists \varphi: \mpar{\fvars{f} \setminus V} \to \values, \vdash f\psubst{\sigma \uplus \varphi}} with \(f \in \formulae, \sigma: V \to \values\).

From there the expression algebra is fixed.
With these common definitions and notations settled, the objects of interest can now be defined.
\begin{defi}[Open automaton]
An open automaton is a tuple \(\OA{S}{i}{V}{\varphi}{J}{T}\) with \(S\) the set of states, \(i \in S\) the initial state, \(V\) the set of variable names unique to this automaton, \(\varphi: V \to \values\) the initial valuation of variables, \(J\) the set of hole names and \(T\) a set of open-transitions.

\(S, V, J\) are arbitrary sets, only \(J\) is required to be finite.
\end{defi}
The variable names may clash when considering two automata, in that case we suppose that we can still distinguish the variables in the formulas.
All the possible open transitions, from which transitions in \(T\) are selected, are defined below.
\begin{defi}[Open transition]
An open transition is a tuple \nmm{\OT{s}{s'}{\alpha\mpar{v_\alpha}}{\mpar{\beta_j\mpar{v_{\beta j}}}^{j \in J'}}{g}{\psi}} with \(s, s' \in S\) the source and target states, \(\alpha \in A\) an action label, \(J' \subseteq J\) the holes involved, \(\beta_j \in A\) the action labels of the holes, \(v_\alpha, v_{\beta j}^{j \in J'}\) fresh distinct variables, \(g \in \formulae\) the guard and \(\psi: V \to \terms\) the new value of the variables.

The set of action labels \(A\) is arbitrary and implicit.
To simplify formulae, the notations \(\alpha \defnotation \alpha\mpar{v_\alpha}\) and \(\alpha = \beta \defnotation \alpha = \beta \wedge v_\alpha = v_\beta\) will be used.
The guard and new values of the variables can use any defined variable: \nmm{\bigcup_{v \in V} \fvars{\psi\mpar{v}} \cup \fvars{g} \subseteq V \cup \mbrc{v_\alpha} \cup \mset{v_{\beta j}}{j \in J'}}.
\end{defi}
Now we can define some utilitary functions:
\begin{defi}[Guard, Variables, Out-transition, In-transition]
\(\mathrm{OT}\mpar{s}\) are called the out-transitions of \(s\) and \(\mathrm{IT}\mpar{s}\) and its in-transitions.
\begin{align*}
	\mathit{guard}\mpar{\OTg} & \defnotation g & \mathit{}
	\fvars{\OT{s}{s'}{\alpha\mpar{v_\alpha}}{\mpar{\beta_j\mpar{v_{\beta j}}}^{j \in J'}}{g}{\psi}} & \defnotation \mbrc{v_\alpha} \cup \mset{v_{\beta j}}{j \in J'} \\
	\mathrm{OT}\mpar{s} & \defnotation \mbrc{\OTg \in T} &
	\mathrm{IT}\mpar{s'} & \defnotation \mbrc{\OTg \in T}
\end{align*}
\end{defi}
The intuition of an open automaton is a partially defined LTS with variables, guards on transitions and parametrised actions.

\section{Semantic and composition of Open Automata} % This is taken from 2007.10770.pdf
% TODO: Intro
When the auomaton is in a state \(s \in S\) with a valuation of its variables \(\sigma \in V \to \values\), it cannot perform transitions \nmm{t \defobject \OTg \in T} such that \(\nvdash g\psubst{\sigma}\).
After performing the transition and emitting action \(\alpha\), the automaton is in the state \(s'\) with valuation \(\psi\psubst{\sigma \uplus \varphi}\) for some \(\varphi: \fvars{t} \to \values\).
The holes are synchronised using a handshake like mecanism when the transition happens.
\begin{exi}
\end{exi}

An open automaton can be partially specified, the partial specification comes from the holes.
A hole can be filled with an open automaton, this operation is called composition.
\begin{defi}[Composition of open automata]
The composition of \(\mathrm{OA}_c \defobject \OA{S_c}{i_c}{V_c}{\varphi_c}{J_c}{T_c}\) in the hole \(k \in J_p\) of \(\mathrm{OA}_p \defobject \OA{S_p}{i_p}{V_p}{\varphi_p}{J_p}{T_p}\) is:
\begin{align*}
	\mathrm{OA}_p\subst{\mathrm{OA}_c}{k} \defnotation & \OA{S_p \times S_c}{\mpar{i_p, i_c}}{V_p \uplus V_c}{\varphi_p \uplus \varphi_c}{J_c \uplus J_p \setminus \mbrc{k}}{T} \\
	\text{With } T \defobject & \mset{\OT{\mpar{s_p, s_c}}{\mpar{s'_p, s'_c}}{\alpha_p}{\beta_j^{j \in J'_c \uplus J'_p \setminus \mbrc{k}}}{g_p \wedge g_c \wedge \alpha_c = \beta_k}{\psi_p \uplus \psi_c}}{\OTx{p}{}{}{p} \in T_p, \OTx{c}{}{}{c} \in T_c} \\
	& \cup \mset{\OT{\mpar{s_p, s_c}}{\mpar{s'_p, s_c}}{\alpha_p}{\beta_j^{j \in J'_p}}{g_p}{\psi_p}}{\OTx{p}{}{}{p}, k \notin J', s_c \in S_c}
\end{align*}
Similarly to expression substitution, the parallel composition is noted \(\mathrm{OA}\psubst{\mathrm{OA}_j^{j \in J}}\).
\end{defi}
The actions emitted when \(\mathrm{OA}_c\) makes a transition is sychronised with the action of the hole \(k\) in transitions of \(\mathrm{OA}_p\) which have it as a hole actions (first transition set, \(\alpha_c = \beta_k\)).
No transition of \(\mathrm{OA}_c\) is performed when a transition that do not refer to the hole \(k\) is performed in \(\mathrm{OA}_p\) (second transition set, \(k \notin J'\)).
The composition may look like a handshake, with both automata running in parallel (product of states and joint variables) but it is actually not symmetric.
\begin{exi}
\end{exi}

@@

The important properties we will consider here are:
\begin{description}
\item[Reflexivity:] \(\forall a, a \leq a\);
\item[Transitivity:] \(\forall a\, b\, c, a \leq b \wedge b \leq c \implies a \leq c\);
\item[Preorder:] Reflexivity and transitivity;
\item[Composition completeness:] \(\forall a\, b, a\mbrk{b} \leq a\);
\item[Composition correctness:] \(\forall a\, b, a \leq b \implies \exists c, a =_{FH} b\mbrk{c}\);
\item[Composition context equivalence:] \(\forall a\, b\, c, a \leq b \implies a\mbrk{c} \leq b\mbrk{c}\);
\item[Composition congruence:] \(\forall a\, b\, c, a \leq b \implies c\mbrk{a} \leq c\mbrk{b}\);
\item[Composition compatibility:] \(\forall a\, b\, c\, d, a \leq b \wedge c \leq d \implies c\mbrk{a} \leq d\mbrk{b}\);
\item[FH-bisimulation compatibility:] \(\forall a\, b\, c, d, a =_{FH} b \wedge c =_{FH} d \wedge a \leq c \implies b \leq d\);
\item[Deadlock reduction:] If \(a \leq b\) then \(a\) does not introduces new deadlocks, noted \(a \sqsubseteq b\) in the following; % TODO: Find a better place to introduce a better definition of deadlock reduction.
\item[Livelock reduction:] If \(a \leq b\) then \(a\) does not introduces new livelocks, noted \(a \preceq b\) in the following; % TODO: Find a better place to introduce a better definition of livelock reduction.
\item[\(\tau\)-stuttering:] .% TODO
\end{description}
``Deadlock reduction" conflicts with ``composition completeness" by disallowing some unwanted cases where an automaton in a hole cannot produce any action that out-transitions expect. % And some other do too
In particular filling a hole with the deadlock automaton (one state without out-transitions) is not considered a refinement for a no deadlock introductive refinement.
``Composition completeness" can be modified into ``Deadlock reduction completeness" to be compatible with ``Deadlock reduction": \(\forall a\, b, a\mbrk{b} \sqsubseteq a \implies a\mbrk{b} \leq a\).
Only the latter will be considered.
Similar issues arise on other properties and can be fixed by changing the prerequisites to insure that the most deadlock-prone composition of automata is composition complete:
\begin{description}
\item[Composition context equivalence modified:] \(\forall a\, b\, c, a \leq b \wedge a\mbrk{c} \leq a \implies a\mbrk{c} \leq b\mbrk{c}\);
\item[Composition congruence modified:] \(\forall a\, b\, c, a \leq b \wedge c\mbrk{a} \leq c \implies c\mbrk{a} \leq c\mbrk{b}\);
\item[Composition compatibility modified:] \(\forall a\, b\, c\, d, a \leq b \wedge c \leq d \wedge c\mbrk{a} \leq c \implies c\mbrk{a} \leq d\mbrk{b}\);
\end{description}

On the other hand there is a property that was discussed, that may arise when designing a refinement relation and that is interesting but actually unwanted:
\begin{description}
\item[Daisy equivalence:] \(a\mbrk{D} \leq a \wedge a \leq a\mbrk{D}\), where \(D\) is the single state automaton that can produce any action.
\end{description}
This property state that composing a hole with an automaton which can do anything is not a strict refinement.
While it might seem to be a natural property because this automaton is the closest to the meaning of a hole in term of automaton, there is actually a difference between the two.
Daisy equivalence allows to refine by filling a hole but also by going the other way, that is creating a hole name and putting its hole action to any transition (and possibly keeping any of the unmodified transitions).
The unwanted behaviour arise for instance when two independant holes are filled, then a hole is created and its actions are put where the two other actions were.
This effectively merges independant holes, or equivalently allows to consider that an automaton can be plug simultaneously into several holes which is not part of the composition semantics.
\begin{exi}
\end{exi}

The relations will have to be at least preorders and simulations to be called refinement simulation.
However a simulation on open automata is more complex than a simulation on a LTS.
\begin{defi}[Simulation refinement on open automata]
A relation \(\leq\) is a simulation refinement if it is a preorder and let \(\mathrm{OA}_1 \defobject \OA{S_1}{i_1}{J_1}{V_1}{\varphi_1}{T_1}\) and \(\mathrm{OA}_2 \defobject \OA{S_2}{i_2}{J_2}{V_2}{\varphi_2}{T_2}\), the following holds:
\begin{multline*}
	\mathrm{OA}_1 \leq \mathrm{OA}_2 \implies \exists R: \mpar{S_1 \times S_2} \to \formulae_{V_1 \uplus V_2}, \varphi_1 \uplus \varphi_2 \vdash R\mpar{i_1, i_2} \\
	\wedge \forall \mpar{s_1, s_2} \in S_1 \times S_2, t_1 \defobject \OTx{1}{}{1}{1} \in T_1, \sigma: \mpar{V_1 \uplus \fvars{t_1} \uplus V_2} \to \values, \\
	\sigma \vdash R\mpar{s_1, s_2} \wedge g_1 \implies \exists \OTx{2}{}{2}{2} \in T_2, \\
	\sigma \vdash \alpha_1 = \alpha_2 \wedge \bigwedge_\subbox{j \in J'_1 \cap J'_2} \beta_{1j} = \beta_{2j} \wedge g_2 \wedge R\mpar{s'_1, s'_2}\psubst{\psi_1 \uplus \psi_2}
\end{multline*}
Which means a relation is a simulation refinement if when two automata are related then there must be a relation that relates initial states and valuations and for every pair of state and valuation of the automaton variables that are related, and every possible transitions with valuation of its free variables from the first automaton, there is a possible transition and its valuation such that the produced action matches, the holes actions with same names match and the target states are related.
\end{defi}
This is a natural extenion of the notion of simulation on LTS, which is the same definition without valuation and holes (and guards).
It will serve as a watchdog for any relation that will be defined.

% Working definition next section

\section{Preliminary refinement relations}
The main objective of the refinement relation in this section is to be able to say that a composition of two automata is a refinement of the base automaton (\(a\mbrk{b} \leq a\)).
However composing an open automaton can give an automaton where not much can be said in terms of hole indicies.
So looking at restricted cases where holes are related in a specific manner can help understanding the general case.
For two open automata \(\mathrm{OA}_1 \defobject \OA{S_1}{i_1}{J_1}{V_1}{\varphi_1}{T_1}\) and \(\mathrm{OA}_2 \defobject \OA{S_2}{i_2}{J_2}{V_2}{\varphi_2}{T_2}\), the following refinement relations are defined.

>>>>

\input{preamble.tex}

@@

Family of values, or equivalently maps will be noted \(\mset{i \mapsto x_i}{i \in I}\), \(\mset{i \gets x_i}{i \in I}\) or \(x_i^{i \in I}\).
The latter will only be used when unambiguous; for instance \(\mpar{ax}^{x \in \setR}\) represents a scaling function, \(c^{i \in I}\) is a constant function over \(I\), but \(\mbrc{\alpha \mapsto 1, \beta \mapsto 2, \gamma \mapsto 3}\) must be represented with one of the two first notation.
The disjoint union of two maps \(\varphi: I \to X\) and \(\psi: J \to Y\) with \(I \cap J = \emptyset\) is \(\varphi \uplus \psi: I \uplus J \to X \cup Y\).

@@

\begin{defi}[Free variables, Closed terms, Closed formulae]
\defitem \(\fvars{e}\) is the set of unbound variables in \(e \in E\).
\defitem The terms restricted to variables in \(V\) are \(\rterms{V} \defnotation \mset{t \in \terms}{\fvars{t} = \emptyset}\); \(\rterms{V} \subset \terms\).
\defitem The formulae restricted to variables in \(V\) are \(\rformulae{V} \defnotation \mset{f \in \formulae}{\fvars{f} = \emptyset}\); \(\rformulae{V} \subset \terms\).
\defitem The closed terms are terms restricted to variables in \(\emptyset\), \(\cterms\).
\defitem The closed formulae are formulae restricted to variables in \(\emptyset\), \(\cformulae\).
\end{defi}
By extension, expressions restricted to variables and closed expressions are also defined.
A closed expression is an expression without unbound variable.
\begin{defi}[Interpretation, Values, Satisfiability, (Parallel) substitution]
We assume that the following are given:
\defitem The values \(\values\), which are interpretations of closed terms;
\defitem The validity relation on closed formulae, \({\vdash} \subseteq \cformulae\);
\defitem The substitution in \(e \in E\) of \(x \in \fvars{e}\) by \(t \in \terms\), \(e\subst{t}{x}\);
\defitem The parallel substitution in \(e \in E\) of variables in \(V\) by \(\psi: V \to \terms\), \(e\psubst{\psi}\).
\end{defi}
For the parallel substitution, the set \(V\) is not required to be a subset of \(\fvars{e}\).
In the case where it is not, the variables in \(V \setminus \fvars{e}\) are not substituted.
The substitutions might give a nonsensical expression; for instance let the terms be integers and pairs with (pointwise) addition, \(\mpar{a + b}\psubst{a \mapsto 7, b \mapsto \mpar{4, 5}}\) is not a valid term.
This can be guarded with the atom \(e\subst{t}{x} \in E\) and \(e\psubst{\psi} \in E\) and it will implicitely be the case, for instance when there is quantification on \(t\) and \(\psi\), to simplify notations.

The interpretation of terms is supposed to be decidable.
The validity of formulae might not be decidable nor complete nor consistent, however we will pretend like they are because these are really hard problems for logicians that we don't want to deal with.\footnote{In practise a term algera with only equality is used, and it is has all these properties.}

A value \(v\) may be used for keeping a variable state, and then injected in terms for substitution.
While this is correct when \(\values \subseteq \cterms\), in the other cases this will be used as a shorthand for substitution with any term which can be interpreted as \(v\).
A quantifier followed by a finite set will be used as a shorthand for the quantification on every variable in the set, for instance \(\forall A, \exists B, P\) with \(A = \mbrc{a_1, \dots, a_\card{A}}, B = \mbrc{b_1, \dots, b_\card{B}}\) means \(\forall a_1, \dots, \forall a_\card{A}, \exists b_1, \dots, \exists b_\card{B}, P\).

The validity relations will be noted using the classical relational notation \(\vdash f \defnotation {f \in {\vdash}}\).
The usual validity of a formula \(f \in \formulae\) is noted \({\vdash f} \defnotation {\vdash \forall \fvars{f}, f}\).
The usual validity of a formula \(f \in \formulae\) with some fixed valuation \(\sigma: V \to \values\) is noted \({\sigma \vdash f} \defnotation {\vdash \forall\mpar{\fvars{f} \setminus V}, f\psubst{\sigma}}\).

From there the expression algebra is fixed.
With these common definitions and notations settled, the objects of interest can now be defined.
\begin{defi}[Open automaton]
A semantic open automaton is a tuple \(\OA{S}{i}{V}{\varphi}{J}{T}\) with \(S\) the set of states, \(i \in S\) the initial state, \(V\) the set of variable names unique to this automaton, \(\varphi: V \to \values\) the initial valuation of variables, \(J\) the set of hole names and \(T\) a set of open-transitions.

\(S, V, J\) are arbitrary sets, only \(J\) is required to be finite.
\end{defi}
The variable names may clash when considering two automata, in that case we suppose that we can still distinguish the variables in the formulas.
All the possible open transitions, from which transitions in \(T\) are selected, are defined below.
\begin{defi}[Open transition]
A semantic open transition is a tuple \nmm{\OT{s}{s'}{\alpha\mpar{v_\alpha}}{\mpar{\beta_j\mpar{v_{\beta j}}}^{j \in J'}}{g}{\psi}} with \(s, s' \in S\) the source and target states, \(\alpha \in \actions\) an action label, \(J' \subseteq J\) the holes involved, \(\beta_j \in \actions\) the action labels of the holes, \(v_\alpha \in \rterms{V}\) the emitted action, \(v_{\beta j}^{j \in J'} \in \rterms{V}\) the hole actions, \(g \in \rformulae{V}\) the guard and \(\psi: V \to \rterms{V}\) the new value of the variables.
\end{defi}
The set of action labels \(\actions\) is arbitrary and implicit.
To simplify formulae, the notations \(\alpha \defnotation \alpha\mpar{v_\alpha}\) and \(\alpha = \beta \defnotation \alpha = \beta \wedge v_\alpha = v_\beta\) will be used.

Now we can define some utilitary functions:
\begin{defi}[Guard, Out-transition]
\(\fOT{s}\) are called the out-transitions of \(s\).
\begin{align*}
	\fguard{\OTg} & \defnotation g &
	\mathit{target}\mpar{\OTg} & \defnotation s' \\
	\mathit{Post}\mpar{\OTg} & \defnotation \psi &
	\fOT{s} & \defnotation \mbrc{\OTg \in T}
\end{align*}
\end{defi}
The intuition of a semantic open automaton is a partially defined LTS with variables, guards on transitions and parametrised actions.
From this point semantic open automata and open transitions will be called automata and transitions for simplicity.

\section{Semantic and composition of Open Automata} % This is taken from 2007.10770.pdf
% TODO: Intro
When the auomaton is in a state \(s \in S\) with a valuation of its variables \(\sigma: V \to \values\), it cannot perform transitions \nmm{\OTg \in T} such that \(\neg g\psubst{\sigma}\).
After performing the transition and emitting action \(\alpha\), the automaton is in the state \(s'\) with valuation \(\psi\psubst{\sigma}\).
The holes are synchronised using a handshake like mecanism when the transition happens.
\begin{exi}
\end{exi}

An open automaton can be partially specified, the partial specification comes from the holes.
A hole can be filled with an open automaton, this operation is called \Quentin{composition}{Or substitution, or filling, what do you prefer?}.
\begin{defi}[Composition of open automata]
The composition of \(A_c \defobject \OA{S_c}{i_c}{V_c}{\varphi_c}{J_c}{T_c}\) in the hole \(k \in J_p\) of \(A_p \defobject \OA{S_p}{i_p}{V_p}{\varphi_p}{J_p}{T_p}\) is
\begin{align*}
	A_p\subst{A_c}{k} \defnotation & \OA{S_p \times S_c}{\mpar{i_p, i_c}}{V_p \uplus V_c}{\varphi_p \uplus \varphi_c}{J_c \uplus J_p \setminus \mbrc{k}}{T} \\
	\text{With } T \defobject & \mset{\OT{\mpar{s_p, s_c}}{\mpar{s'_p, s'_c}}{\alpha_p}{\beta_j^{j \in J'_c \uplus J'_p \setminus \mbrc{k}}}{g_p \wedge g_c \wedge \alpha_c = \beta_k}{\psi_p \uplus \psi_c}}{\OTx{p}{}{}{p} \in T_p, \OTx{c}{}{}{c} \in T_c} \\
	& \cup \mset{\OT{\mpar{s_p, s_c}}{\mpar{s'_p, s_c}}{\alpha_p}{\beta_j^{j \in J'_p}}{g_p}{\psi_p}}{\OTx{p}{}{}{p}, k \notin J', s_c \in S_c}
\end{align*}
Similarly to expression substitution, the parallel composition is noted \(A\psubst{A_j^{j \in J}}\).
\end{defi}
The actions emitted when \(A_c\) makes a transition is sychronised with the action of the hole \(k\) in transitions of \(A_p\) which have it as a hole actions (first transition set, \(\alpha_c = \beta_k\)).
No transition of \(A_c\) is performed when a transition that do not refer to the hole \(k\) is performed in \(A_p\) (second transition set, \(k \notin J'\)).
The composition may look like a handshake, with both automata running in parallel (product of states and joint variables) but it is actually not symmetric.
\begin{exi}
\end{exi}

@@

The important properties we will consider here are:
\begin{defi} A relation \(\leq\) is
\defitem \textbf{reflexive} iff \(\forall a, a \leq a\);
\defitem \textbf{transitive} iff \(\forall a\, b\, c, a \leq b \wedge b \leq c \implies a \leq c\);
\defitem \textbf{a preorder} iff it is reflexive and transitive;
\defitem \textbf{complete w.r.t.\@ composition} iff \(\forall a\, b, a\mbrk{b} \leq a\);
\defitem \textbf{correct w.r.t.\@ composition} iff \(\forall a\, b, a \leq b \implies \exists c, a \overset{FH}= b\mbrk{c}\);
\defitem \textbf{context equivalent for composition} iff \(\forall a\, b\, c, a \leq b \implies a\mbrk{c} \leq b\mbrk{c}\);
\defitem \textbf{congruent for composition} iff \(\forall a\, b\, c, a \leq b \implies c\mbrk{a} \leq c\mbrk{b}\);
\defitem \textbf{compatible with composition} iff \(\forall a\, b\, c\, d, a \leq b \wedge c \leq d \implies c\mbrk{a} \leq d\mbrk{b}\);
\defitem \textbf{compatibile with FH-bisimulation} iff \(\forall a\, b\, c, d, a \overset{FH}= b \wedge c \overset{FH}= d \wedge a \leq c \implies b \leq d\).
\end{defi}
Reflexivity, transitivity and preorder are classical properties on relations.
Completeness w.r.t.\@ composition is the fact that every composition is considered a refinement.
Correctness w.r.t.\@ composition is the fact that a refinement correspond to the left automaton being equivalent to some composition of the right automaton.
Context equivalence is the refinement being compatible with composition with the same automaton.
Congruence is the refinement being compatible with being composed in the same automaton.
Compatibility with composition is the two latter at the same time.
Finally compatibility with FH-bisimulation is the refinement not being able to distinguish two FH-bisimilar automata.

The relations will have to be at least preorders, simulations and forbid the introduction of deadlocks (state where no transition is possible) to be called refinement simulation.
On top of that weak refinement are refinement where some actions, typically one named \(\tau\), are considered not observable and do not have to be matched.
In the weak refinement the relations have to forbid the introduction of livelocks (only a sequence of \(\tau\) possible) and respect \(\tau\)-stuttering \marginpar{cite many things arround here}.
While being preorder is simple to express for a relation, being a simulation, do not introduce new deadlocks/livelocks and \(\tau\)-stuttering are properties about the states of the related automata, therefore they are more complex to express.
We will need a preliminary concept before defining them.
\begin{defi}[Relation under predicate]
A relation under predicate between states of two automata \(\OA{S_1}{i_1}{J_1}{V_1}{\varphi_1}{T_1}\) and \(\OA{S_2}{i_2}{J_2}{V_2}{\varphi_2}{T_2}\) is a function in \(S_1 \times S_2 \to \rformulae{V_1 \uplus V_2}\).
Two states \(s_1 \in S_1, s_2 \in S_2\) are related under a valuation \(\sigma: V_1 \uplus V_2 \to \values\) iff \(\sigma \vdash R\mpar{s_1, s_2}\).
\end{defi}
Now let's adapt the properties about states of automata to open automata.
\begin{defi}[Simulation on open automata]
A relation \(\leq\) is a simulation iff for any two related automata \(\OA{S_1}{i_1}{J_1}{V_1}{\varphi_1}{T_1} \leq \OA{S_2}{i_2}{J_2}{V_2}{\varphi_2}{T_2}\), there is a relation under predicate between their states \(R: S_1 \times S_2 \to \rformulae{V_1 \uplus V_2}\) that satisfies both
\defitem Initial states are related under initial valuations: \(\varphi_1 \uplus \varphi_2 \vdash R\mpar{i_1, i_2}\);
\defitem From related states, all out-transitions from \(A_1\) can be simulated in \(A_2\) and their target states are related:
\begin{multline*}
	\forall \mpar{s_1, s_2} \in S_1 \times S_2, \sigma: V_1 \uplus V_2 \to \values, \\
	\bigsymb{\forall} \OTx{1}{}{1}{1} \in T_1, \mpar{\sigma \vdash R\mpar{s_1, s_2} \wedge g_1} \implies \bigsymb{\exists} \OTx{2}{}{2}{2} \in T_2, \\
	\sigma \vdash \alpha_1 = \alpha_2 \wedge \bigwedge_\subbox{j \in J'_1 \cap J'_2} \beta_{1j} = \beta_{2j} \wedge g_2 \wedge R\mpar{s'_1, s'_2}\psubst{\psi_1 \uplus \psi_2}
\end{multline*}
\end{defi}
The last formula means that for every pair of related states under valuation of the automaton variables and every possible transitions (\(\sigma \vdash g_1\)) from the first automaton, there is a possible transition (\(\sigma \vdash g_2\)) such that the produced action matches (\(\alpha_1 = \alpha_2\)), the holes actions from same hole names (\(J'_1 \cap J'_2\)) match (\(\beta_{1j} = \beta_{2j}\)) and the target states are related after variable update.
This is a natural extenion of the notion of simulation on LTS, which is the same definition without variables and holes and guards.
It will serve as a watchdog for any relation that will be defined with ``simulation" in its name, except weak-simulation where the transition from the first automaton that produce \(\tau\) actions are excluded.

\begin{defi}[Deadlock reduction]
A simulation \(\leq\) is deadlock reducing if for all related automata \(A_1 \leq A_2\), where \(A_1 \defobject \OA{S_1}{i_1}{V_1}{\varphi_1}{J_1}{T_1}\) and \(A_2 \defobject \OA{S_2}{i_2}{V_2}{\varphi_2}{J_2}{T_2}\), there is a witness \(R: S_1 \times S_2 \to \rformulae{V_1 \uplus V_2}\) of the fact that it is a simulation that also satisfies
\begin{multline*}
	\forall \mpar{s_1, s_2} \in S_1 \times S_2, \sigma: V_1 \uplus V_2 \to \values, \\
	\mpar{\exists t_2 \in \fOT{s_2}, \sigma \vdash R\mpar{s_1, s_2} \wedge \fguard{t_2}} \implies \bigsymb{\exists} \OTx{2}{}{2}{2} \in T_2, \OTx{1}{}{1}{1} \in T_1, \\
	\sigma \vdash g_2 \wedge \alpha_1 = \alpha_2 \wedge \bigwedge_\subbox{j \in J'_1 \cap J'_2} \beta_{1j} = \beta_{2j} \wedge g_1 \wedge R\mpar{s'_1, s'_2}\psubst{\psi_1 \uplus \psi_2}
\end{multline*}
\end{defi}
This definition extends the requirements of a simulation by requiring from the relation that on related states, if there is at least one possible transition in the second automata then there must be a matching possible transition in the first.
This requirement effectively prevents the presence of deadlock state in the first automaton if there is a related non-deadlock state in the second.

% TODO LATER: livelock and tau-stuttering
% \item[Livelock reduction:] If \(a \leq b\) then \(a\) does not introduces new livelocks, noted \(a \preceq b\) in the following; % TODO: Find a better place to introduce a better definition of livelock reduction.
% \defitem \textbf{\(\tau\)-stuttering} iff . % TODO
